---
title: "Stat556: Causal Modeling Project"
author: "Bhuvan, Darshan, Yash"
date: "March 15, 2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library("pcalg")
library("graph")
library("Rgraphviz")
library(abind)
library(corpcor)
library(sfsmisc)
library(robustbase)
library("dplyr")
source("http://www.stat.washington.edu/tsr/s566/labs/y0y1polytopenew-rgl-col.R")
library(rgl)

plotcpdag <- "Rgraphviz" %in% print(.packages(lib.loc = .libPaths()[1]))
```


```{r}
data = read.csv("../data/data_filtered.csv")
full_data = read.csv("../data/raw_data.csv")

alpha = 0.05
```

# EDA
```{r}
boxplot(data$delta5)
summary(data$delta5)
plot(data$delta5)
hist(data$delta5)
sum(data$delta5 < 0) / length(data$delta5)
```

```{r}
# Function for Neyman's ACE analysis.
get_neyman_ace = function(trt_data, control_data, alpha) {
  neyman_ace = mean(trt_data) - mean(control_data)
  var_neyman_ace = var(trt_data)/length(trt_data) + var(control_data)/length(control_data)

  ci_lower = neyman_ace - qnorm(1-alpha/2) * sqrt(var_neyman_ace)
  ci_upper = neyman_ace + qnorm(1-alpha/2) * sqrt(var_neyman_ace)
  
  ace_ci = c(ci_lower, ci_upper)
  
  cat("Estimate of Neyman ACE:", neyman_ace, "\n")
  cat("Estimate of Variance of ACE:", var_neyman_ace, "\n")
  cat("COnfidence Interval: (", ci_lower, ",", ci_upper, ")\n")
}

# Function to get CPDAG and PAG.
get_dags  = function(mydata, alpha) {
  # Graphical display of correlation matrix.
  pairs(mydata, lower.panel = NULL)

  names <- attr(mydata, "names")
  n <- nrow(mydata)
  p <- ncol(mydata)
  indepTest <- gaussCItest
  suffStat <- list(C=cor(mydata), n = n)
  
  ## Estimate CPDAG
  pc.fit <- pc(suffStat, indepTest=indepTest, p = p, alpha = alpha, verbose = TRUE)
  showAmat(pc.fit)
  showEdgeList(pc.fit, names)
  
  print(names)
  
  if (plotcpdag) {
    plot(pc.fit, main = "Estimated CPDAG",labels=names)
  }
  
  ##### Using the FCI Algorithm to estimate an ancestral graph
  ## Estimate Partial Ancestral Graph (PAG)
  fci.fit <- fci(suffStat, indepTest, p = p, labels=names, alpha = alpha, verbose = TRUE)
  
  print(fci.fit@amat)
  
  if (plotcpdag) {
    dev.off()
    plot(fci.fit)
  }
  
  print(fci.fit@sepset)
}
```


## Q1: Neyman's method to find ACE of Acupunture on HSS.

```{r}
# For delta5 continuous.
control_data = subset(data, group==0)
trt_data = subset(data, group==1)
get_neyman_ace(trt_data$delta5, control_data$delta5, alpha=alpha)
```

## Q2: Adjust for confounders and find ACE.
- Include all baseline confounders and get a PAG.
- Get a SWIG from PAG for X=x*
- Analyse d-separations and identify minimal set of confounders.
- Backdoor formula/Matching algorithm to find ACE.

```{r}
# For delta5 continuous:
baseline_covariates_xy = select(data, group, age, sex, migraine, chronicity, pk1, f1, painmedspk1, prophmqs1, delta5)
get_dags(baseline_covariates_xy, alpha)
```

Since intervention is done on X=group, and our response variable is Y=delta5 (change in HSS from baseline and 12-months), when SWIG G* is written for every graph in the above PAG, we find that X=group is d-separated from Y=delta5 unconditionally. Therefore, from faithfulness assumption (is it required?), we find that X=group is independent of Y=delta5.

Therefore, the ACE observed in Q1, is valid because we do not need to account for confounders using backdoor formula.

## Q3: Complier analysis by IV model.
Binarizing the response variable (delta5), to check if headache imporved or worsened.

```{r}
# Find Neyman ACE for delta5 binarized.
data$delta5_binary = ifelse(data$delta5 > 0, 1, 0)
get_neyman_ace(trt_data$delta5_binary, control_data$delta5_binary, alpha=alpha)
```

```{r}
# Check SWIG for delta5 binarized.
baseline_covariates_xy_binary = select(data, group, age, sex, migraine, chronicity, pk1, f1, painmedspk1, prophmqs1, delta5_binary)
get_dags(baseline_covariates_xy_binary, alpha)
```

```{r}
# IV Model: Considering delta5 binary.
data$acuptreatments[is.na(data$acuptreatments)] = 0
median_acuptreatments = median(data$acuptreatments, na.rm=T)
data$complier = ifelse(data$acuptreatments >= median_acuptreatments, 1, 0)

z0_data = subset(data, group==0)
z1_data = subset(data, group==1)
# p(y0, x0 | z0), p(y0, x1 | z0), p(y1, x0 | z0), p(y1, x1 | z0), p(y0, x0 | z1), p(y0, x1 | z1), p(y1, x0 | z1), p(y1, x1 | z1)
iv_data = c(
  nrow(subset(z0_data, delta5_binary==0 & complier==0)) / nrow(z0_data), # p(y0, x0 | z0)
  nrow(subset(z0_data, delta5_binary==0 & complier==1)) / nrow(z0_data), # p(y0, x1 | z0)
  nrow(subset(z0_data, delta5_binary==1 & complier==0)) / nrow(z0_data), # p(y1, x0 | z0)
  nrow(subset(z0_data, delta5_binary==1 & complier==1)) / nrow(z0_data), # p(y1, x1 | z0)
  nrow(subset(z1_data, delta5_binary==0 & complier==0)) / nrow(z1_data), # p(y0, x0 | z1)
  nrow(subset(z1_data, delta5_binary==0 & complier==1)) / nrow(z1_data), # p(y0, x1 | z1)
  nrow(subset(z1_data, delta5_binary==1 & complier==0)) / nrow(z1_data), # p(y1, x0 | z1)
  nrow(subset(z1_data, delta5_binary==1 & complier==1)) / nrow(z1_data)  # p(y1, x1 | z1)
)

check.iv.ineqs(iv_data)

iv_data_z0 = iv_data[1:4]
simp <- do.simplex(phi=30,theta=120,r=1000,main="IV Data; 3D view")
do.polytope(iv_data_z0, simp)

iv_data_z1 = iv_data[5:8]
simp <- do.simplex(phi=30,theta=120,r=1000,main="IV Data; 3D view")
do.polytope(iv_data_z1, simp)

simp <- do.simplex(phi=30,theta=120,r=1000,main="IV Data; 3D view")
do.polytope(iv_data, simp)

ace.bnds <- ace.bounds(iv_data)
cat("ACE Bounds:", ace.bnds)

simp <- do.simplex(phi=-90,theta=90,r=1000,main="IV Data; 2D view")
do.polytope(iv_data, simp)

do.ace.line(ace.bnds[1], simp) # lower bound
do.ace.line(ace.bnds[2], simp) # upper bound
```

## Q4: Mediation Analysis of Prophylactic Medication MQS score.
```{r}
mediation_mqs_data = select(data, group, prophmqs5, delta5)
get_dags(mediation_mqs_data, alpha)
```

From the SWIG that follows (for all the DAGs that the PAG represents), we find that delta5 is d-sep from prophmqs5, suggesting that mediation effect is unlikely.

## Q5: Dropout Analysis: 1. baseline confounders, 2. 3-month HSS analysis.
```{r}
stay_data = subset(full_data, !is.na(full_data$delta5) & !is.na(full_data$delta2))
dropout_data = subset(full_data, is.na(full_data$delta5) & !is.na(full_data$delta2))

# Analyse the difference in baseline covariates: age, sex, migraine, chronicity, pk1, f1, painmedspk1, prophmqs1
baseline_covariates = c("age", "sex", "migraine", "chronicity", "pk1", "f1", "painmedspk1", "prophmqs1")
p_value_list = c()
reject_list = c()

for (i in 1:length(baseline_covariates)) {
  v = baseline_covariates[i]
  t_test = t.test(stay_data[v], dropout_data[v], var.equal = FALSE)
  p_value_list[i] = t_test$p.value
  reject_list[i] = t_test$p.value < alpha
}

summary_df = data.frame(baseline_covariate=baseline_covariates, p_value=p_value_list, reject_H0=reject_list)
print(summary_df)
```
Limitation: We ignore those patients who did not give data at 3-months but had given data at 12-months. There are only 6 such patients.

We observe that there is no significant difference between any of the baseline covariate values the people who dropped-out of the study after 3-months.

```{r}
# Compare the delta2 values between these two groups.
t_test = t.test(subset(stay_data, group==1)$delta2, subset(dropout_data, group==1)$delta2, var.equal = FALSE)
t_test$p.value
cat("P-Value:", t_test$p.value, "\n")
cat("Reject H0:", t_test$p.value <= 0.05, "\n")
```
No evidence for significant difference of delta2 between people who dropped out and those who stayed in till 12-months.

## Q6: SF-36 health status, 9-aspects Causal DAG. 1. separate 2. including X.
```{r}
filtered_data = subset(data, !is.na(data$pf5))

# For SH-36 Health status:
filtered_data$pf5_delta  = filtered_data$pf5  - filtered_data$pf1
filtered_data$rlp5_delta = filtered_data$rlp5 - filtered_data$rlp1
filtered_data$rle5_delta = filtered_data$rle5 - filtered_data$rle1
filtered_data$ef5_delta  = filtered_data$ef5  - filtered_data$ef1
filtered_data$ewb5_delta = filtered_data$ewb5 - filtered_data$ewb1
filtered_data$sf5_delta  = filtered_data$sf5  - filtered_data$sf1
filtered_data$p5_delta   = filtered_data$p5   - filtered_data$p1
filtered_data$gen5_delta = filtered_data$gen5 - filtered_data$gen1
filtered_data$hc5_delta  = filtered_data$hc5  - filtered_data$hc1

sf36_health_aspects_x = select(filtered_data, group, pf5_delta, rlp5_delta, rle5_delta, ef5_delta, ewb5_delta, sf5_delta, p5_delta, gen5_delta, hc5_delta)

get_dags(sf36_health_aspects_x, alpha)
```

Looks like the treatment group assignment is independent of most aspects of the SF-36 health status.

Emotional well-being (ewb), general health (gen) and energy fatigue (ef) are interdependent and group has a causal effect on energy fatigue.

## Q7: Causal DAG for outcome variables.
