---
title: "Analysis of the effect of prophylactic medication (MQS score) and acupuncture treatment on headache"
author: "Darshan, Bhuvan"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(sandwich)
```

```{r}
data = read.csv("../data/data_filtered.csv")
```

#### Problem Description:
We found acupuncture treatment to have a significant effect on the change in Headache Severity Score (HSS) score after adjustment of the confounders. From the study, both the control and treatment group were given a commonly used headache prevention medicine (Prophylactic Medication - measured using MQS score). The treatment group was additionally given acupunture treatments. We suspect that the Prophylactic medication could act as a potential mediator or moderator. Therefore, we test for significant mediation (full or partial). If we do not find enough evidence for mediation, we shall test for moderation (interaction) effect.

#### Statistical Analysis:
We consider the treatment group assignment as primary predictor (X). This is referred as 'Acupunture' since it is a binary treatment group assignment. Our primary ourcome variable is the change in the HSS (Y) after 12-months, i.e., difference between 12-months and baseline. The HSS at 12-months is an aggregate of everyday HSS entered for four weeks by patients after 12-months. During this time, the patients in both, control and treatment group, continued to take the Prophylactic medication quantified by the MQS score (M) referred as MEDS from now onwards.

We need to determine whether MEDS mediated or moderated the effect of Acupunture on the change in the HSS after 12-months. The significance level (alpha) is set at 0.05 for this analysis.

_Note:_ For this analysis, we shall filter the data and consider only patients who completed the study till 12-months (data_filtered.csv).

Mediation Effect Analysis was carried out as follows [4]:
Please refer to the graph from Fig. 1 (#TODO: Need to be put up in report).
X: Treatment group assignment - primary predictor
M: Prophylactic Medication MQS score - Potential mediator
Y: Change in HSS score after 12-months - Response variable

_Note_: The assumptions and the validity of each test needs to be verified.
_Note_: All the below tests need to be conducted with adjustment for the confounders. # TODO: Is it valid to do this? Or would the interpretations change? Check this out: https://stats.stackexchange.com/questions/62247/how-can-i-test-control-variables-within-a-mediation-analysis, which refers to [5]. Can you please check if this way of including confounders is valid (We can ask Thomas also). If not, we can do without confounders, but mention this as limitation.


path a: X -> M
path b: M -> Y
path c: X -> Y

1. Step 1: Conduct a simple regression analysis with X predicting Y to test for path c alone, $Y = B_0 + B_1X + e$. We test the Null Hypothesis that there is no effect of X on Y i.e., $H_0: B_1= 0$.

2. Step 2: Conduct a simple regression analysis with X predicting M to test for path c alone, $M = B_0 + B_1X + e$. We test the Null Hypothesis that there is no effect of X on M i.e., $H_0: B_1= 0$.

3. Step 3: Conduct a simple regression analysis with M predicting Y to test for path c alone, $Y = B_0 + B_1M + e$. We test the Null Hypothesis that there is no effect of M on Y i.e., $H_0: B_1= 0$.

- The steps 1-3 establish the existence of zero-order relationships. If one or more of these relationships are nonsignificant, we decide to conclude that mediation is not possible or likely (although this is not always true [5]). If significant relationships are found from Step 1-3, then we conduct Step 4.

4. Step 4: Conduct a multiple regression analysis with X and M predicting Y, $Y = B_0 + B_1X + B_2M + e$. We test the Null Hypothesis that there is no effect of M on Y when adjusted for X i.e., $H_0: B_2= 0$.

- In Step 4, if the effect of M (path b) remains significant after controlling for X, then some form of mediation is supported. If this is true, and if X is no longer significant when M is controlled, then the finding supports full mediation. And, if X is still significant, the finding supports partial mediation. In order to find the statistical significance of mediation effect, proceed to Step 5 and ignore Step 6.

- In Step 4, if the effect of M (path b) is not significant after controlling for X, then Mediation effect is not likely. Hence, proceed to Step 6 to test for Moderation effect.

5. Step 5: Determine the significance of indirect effect using Sobel Product of Coefficients Approach - $B_i = (B_2)(B_1)$. Here $B_2$ is taken from model from Step 4, and $B_1$ is taken from model from Step 2. Determine the significance using Bootstrap method - bootstrap sampling is done on patient ids, and corresponding values for X, M and Y are collected and $B_i$ is calculated for 5000 bootstrap samples. Bootstrap standard error ($SE(B_i)$) is calculated by taking standard deviation of collected array of $B_i$s. Then the Z statistic - $B_i/SE(B_i)$ is calculated and compared to a Normal distribution to get a p-value. If there exists statistical significance for the indirect effect, then the mediation effect can be considered statistically signifianct. (#TODO: Can we do bootstrap in one place and robust SE in another? Or should we use the same approach in both the places?)

6. Step 6: Conduct a regression analysis with X predicting Y including an interaction term for X:M, $Y = B_0 + B_1X + B_2M + B_3(X)(M) + e$. We test the Null Hypothesis that there is no interaction effect X:M i.e., $H_0: B_3= 0$. This determines the evidence for existence of moderation (interaction) effect of M on X.

```{r}
# counfounders = c(age, sex, chronicity, acupuncturist, prophmqs1, migraine, pk1)

# Checking zero-order relationship: X -> Y
print("Checking zero-order relationship: X -> Y")
s1_model = lm(delta5 ~ group + age + sex + chronicity + acupuncturist + prophmqs1 + migraine + pk1, data=data, )
summary(s1_model)

# Checking zero-order relationship: X -> M
print("Checking zero-order relationship: X -> M")
s2_model = lm(prophmqs5 ~ group + age + sex + chronicity + acupuncturist + prophmqs1 + migraine + pk1, data=data)
summary(s2_model)

# Checking zero-order relationship: M -> Y
print("Checking zero-order relationship: M -> Y")
s3_model = lm(delta5 ~ prophmqs5 + age + sex + chronicity + acupuncturist + prophmqs1 + migraine + pk1, data=data)
summary(s3_model)

# Checking effect of M controlling for X
print("Checking effect of M controlling for X")
s4_model = lm(delta5 ~ prophmqs5 + group + age + sex + chronicity + acupuncturist + prophmqs1 + migraine + pk1, data=data)
summary(s4_model)

# Checking Moderation effect caused by M (interaction effect between X and M)
print("Checking Moderation effect caused by M (interaction effect between X and M)")
moderator_model = lm(delta5 ~ group*prophmqs5 + age + sex + chronicity + acupuncturist + prophmqs1 + migraine + pk1, data=data)
summary(moderator_model)
```

Results and Discussion:
# This can be ignored. Old results, just kept for reference.
1. Mediation effect: It is unlikely for M to have a mediation effect as there is no zero-order significant effect of M on Y. Also, when controlled for X, the effect of M on Y is not significant.

2. Moderator effect: Since we found that M cannot be a mediator, we now consider if M is a Moderator. Considering the interaction model, we see that there is no significant interaction effect between X and M. Therefore, there is not enough evidence to reject the hypothesis that the effect of X on Y doesn't change for unit difference in M.


However, before we pass the judgement, we need to make sure that the assumptions of the test used above have been met. More specifically, we need to test the following for each of the models in the above snippet:

1. Independence
2. Linearity
3. Constant variance of the error terms
4. Normality or Large Sample size

While independence can only be tested from background knowledge of the data collection and experiment design mechanism, for the sake of this analysis, we assume that independence holds and test for the rest.

```{r}
check_assumptions = function(model){
  par(mfrow = c(1,2))
  plot(model, which=c(1:2))
}
```

```{r checkAssumptions: X->Y with confounders}
check_assumptions(s1_model)
```

From the Residuals vs. fitted values plot, we see that linearity and constant variance assumption is met. From the Q-Q plot, we see very minor deviations from normality around the tails which isn't sufficient evidence to rule out normality and since the sample size seems large enough (301), we can say that normality assumption is met. 

```{r checkAssumptions: X->M with confounders}
check_assumptions(s2_model)
```

From the Residuals vs. fitted values plot, we see that linearity and constant variance assumption is met. From the Q-Q plot, we see very minor deviations from normality around the tails which isn't sufficient evidence to rule out normality and since the sample size seems large enough (301), we can say that normality assumption is met. 

```{r checkAssumptions: M->Y with confounders}
check_assumptions(s3_model)
```

From the Residuals vs. fitted values plot, we see that linearity and constant variance assumption is met. From the Q-Q plot, we see very minor deviations from normality around the tails which isn't sufficient evidence to rule out normality and since the sample size seems large enough (301), we can say that normality assumption is met. 

```{r checkAssumptions: M->Y (X controlled) with confounders}
check_assumptions(s4_model)
```

From the Residuals vs. fitted values plot, we see that linearity and constant variance assumption is met. From the Q-Q plot, we see very minor deviations from normality around the tails which isn't sufficient evidence to rule out normality and since the sample size seems large enough (301), we can say that normality assumption is met. 

```{r checkAssumptions: X*M->Y with confounders}
check_assumptions(moderator_model)
```

From the Residuals vs. fitted values plot, we see that linearity and constant variance assumption is met. From the Q-Q plot, we see very minor deviations from normality around the tails which isn't sufficient evidence to rule out normality and since the sample size seems large enough (301), we can say that normality assumption is met. 

#### Limitation:
- We have not explored how confounders could affect the mediation analysis. There are complex methods to test mediation with control variables [5].
- There could be multiple variables which could be potential mediators like Pain medication MQS score, Number of General Practitioner Visits etc. which have not been considered in this analysis. There exists complex procedures to perform meidation analysis for multiple potential mediators [6], which can be included in the future research.

#### Without counfounders test:
```{r}
# counfounders = c(age, sex, chronicity, acupuncturist, prophmqs1, migraine,
# pk1)

# Checking zero-order relationship: X -> Y
print("Checking zero-order relationship: X -> Y")
s1_model = lm(delta5 ~ group, data=data)
summary(s1_model)

# Checking zero-order relationship: X -> M
print("Checking zero-order relationship: X -> M")
s2_model = lm(prophmqs5 ~ group, data=data)
summary(s2_model)

# Checking zero-order relationship: M -> Y
print("Checking zero-order relationship: M -> Y")
s3_model = lm(delta5 ~ prophmqs5, data=data)
summary(s3_model)

# Checking effect of M controlling for X
print("Checking effect of M controlling for X")
s4_model = lm(delta5 ~ prophmqs5 + group, data=data)
summary(s4_model)

# Checking Moderation effect caused by M (interaction effect between X and M)
print("Checking Moderation effect caused by M (interaction effect between X and M)")
moderator_model = lm(delta5 ~ group*prophmqs5, data=data)
summary(moderator_model)
```

#### References:
[4] Web Article - [http://web.pdx.edu/~newsomj/semclass/ho_mediation.pdf](http://web.pdx.edu/~newsomj/semclass/ho_mediation.pdf)

[5]. MacKinnon, David P., Amanda J. Fairchild, and Matthew S. Fritz. "Mediation analysis." Annu. Rev. Psychol. 58 (2007): 593-614.

[6]. VanderWeele, Tyler, and Stijn Vansteelandt. "Mediation analysis with multiple mediators." Epidemiologic methods 2.1 (2014): 95-115.
